Determining sequence-dependent DNA oligonucleotide hybridization and dehybridization mechanisms using coarse-grained molecular simulation, Markov state models, and infrared spectroscopy

### 2021-07-29 cover_letter/response_letter.tex

\end{letter}

%----------------------------------------------------------------------------------------

\clearpage
\newpage

\noteg{Please go through and (1) make sure everything reproduced here is exactly identical to that in the main text and SI (e.g., no formatting codes got messed up and all figures have the correct names) -- please do this carefully (I know it's a pain) since if a reviewer notices that we made a different modification in the manuscript from that which we report in the response letter it puts all of our credibility in question, (2) remove all notes, placeholders, and colored ink where not being used to indicate a reviewer comment (blue) or new text in the revised manuscript (red) -- you next GitHub push should be ready to submit, (3) update all page references to be correct and remove the red ink (e.g., on p.~17 we have added..., (4) reorder and relabel the SI figures such that they are referred to in order as you read through the main text (e.g., right now I think we mention Fig.~S7 before Fig.~S1), (5) clean up the GitHub archive so the cover letter and response letter are in separate directories, old figures have been deleted, and all new figures have the correct names and are < 1 MB in file size.}

@@

\textbf{Author Reply:}   We thank the reviewer for their close reading of the manuscript and are pleased that they find the work impactful and suitable for publication in JACS.

\textcolor{blue}{Major point.
1)      Many of the core conclusions about the hybridization and dehybridzation mechanisms are based on the designation of the relevant metastable states as H, 5S2, 3S2, 5S4, F4, and D. While the authors do an excellent job of demonstrating that 2-6 state models provide an accurate description of the dynamics, in the current form, the paper does not provide sufficient evidence to support their chosen representative microstates and associated nomenclature (the only evidence being the ``representative'' structures shown in Figures 2A, C). A given macrostate represents an ensemble of microstates and it should be demonstrated that a significant fraction of the microstates that have a high probability of belonging to a particular macrostate share the same or similar base-pairing pattern. This would better justify how the authors chose to describe these states as 5S2, 3S2, etc., and provide the reader with more confidence that the mechanisms follow from the simulations. At the least, the authors should describe how the representative microstates (depicted in Figure 2A, C) were obtained.
}

\textbf{Author Reply:} The designation and structural representation of the macrostates is critical to the conclusions of the manuscript and we concur that we should better expose for the reader how the representative microstates were selected, how representative they are of the microstate ensemble within the macrostate, and our analysis of the base-pairing patterns within each macrostate that motivated our naming scheme. We have addressed this by modifying the main text and conducting additional analyses of our microstate and macrostate distributions that we detail in a new figure in the Supporting Information.

@@

%\noteg{I think this comment is really getting at why the cartoon representations are appropriate to describe the state as a whole and is less concerned with how we obtained the molecular rendering. I've tried to address both aspects below and can add more detail the first figure or choose a different kind of representation.}

On \rood{p.~17}, we have added the following text:

``...each of these macrostates. \rood{The representative microstates emblematic of each macrostate were selected randomly from those possessing high ($>$99\%) macrostate membership probabilities. Since the microstate ensemble exhibits a range of conformations, however, it is useful to characterize the degree of structural heterogeneity to determine the degree to which these microstates are emblematic of the distribution. In Fig.~S{\blauw{X}} we present a projection of our macrostates into two physical order parameters -- the degree of 3$^\prime$ shift and degree of 5$^\prime$ shift -- to provide an interpretable embedding of the macrostates that exposes base pairing patterns and structural heterogeneity within the microstate ensemble. In all cases, we find these microstate distributions to be relatively narrowly focussed within the low-free energy core of each macrostate such that a single archetypal microstate is indeed a good representative for the ensemble and an accurate representation of the base pairing pattern of the macrostate.} In Fig. 2b we present...''

%\noter{Fig. SX should contain four panels: (a) Explication of the two order parameters. (b) Current macrostate\_colors\_FES but with improved schematics on x and y axes. Use $d_{3^\prime}^{-1}$ and $d_{5^\prime}^{-1}$ as axis labels. Please also provide a numerical scale for the free energy color map. Perhaps also draw a contour line around each free energy well showing the region that contains, say, 90\% of the probability mass -- the idea is that we want to show that the microstate ensembles are very much contained within a tight distribution at the FE minima. (c) Same as (b) but showing and heat map showing not FE but state membership probabilities. (d) Table listing mean and standard deviation for 3' shift and 5' shift for each state in each system in both (i) real units and (ii) reciprocal units, and the last column showing a superposition of $\sim$50 representative configurations with $>$90\% membership probabilities to try to show the structural variation in the ensemble.}

We also added the new Fig.~S\rood{X} to the Supporting Information where we discuss at some length within the caption the important issues raised by the reviewer. The figure and its caption are reproduced below. \noteg{Please make sure what is reproduced here is exactly the same as that in the SI. Right now the images are different.}

%\noter{When reproducing a main text or SI figure in the response letter, I think it's a good practice to typically (i) reproduce the whole figure and caption and (ii) give it the same numbering as in the manuscript rather than an independent numbering just for the response letter. Figures generated just for the response letter can have some other numbering scheme (e.g., R1, R2, ...) or left without numbers.}

@@

\begin{figure}[ht!]
	\begin{center}
        %\includegraphics[width=150mm]{./revision_figures/macrostate_colors_FES.png}
        \caption*{Figure S\rood{X}. \rood{Macrostate interpretation and structural analysis. The Markov state model macrostates are learned from the simulation trajectories using the featurization and deep learning procedure described in Section 2.1.2 of the main text, but it is informative to perform a \textit{post hoc} projection of the simulation data into a low-dimensional space spanned by physically interpretable order parameters in order to determine the structural characteristics of each macrostate and the structural heterogeneity of the microstate ensemble. (a) Definition of two physically-interpretable order parameters useful in characterizing the structure and base pairing of each microstate under both shifting and fraying. We define the $3^\prime$ shift for a particular base on a strand as the linear distance between that base and the base on the complementary strand offset by two base pairs in the $3^\prime$ direction. This distance reaches a minimum when the two strands are mutually shifted two base pairs in the $3^\prime$ direction. We define $d_{3^\prime}$ as the mean $3^\prime$ shift averaged over the two central bases, represented by red arrows, in the complementary strands. It is more convenient to work the reciprocal $d_{3^\prime}^{-1}$, since it has a finite range whereas $d_{3^\prime}$ diverges for dehybridized strands. We define $d_{5^\prime}^{-1}$ analogously as the reciprocal mean $5^\prime$ shift. We have found this pair of variables extremely informative in revealing the microstate and macrostate structure and base pairing patterns. (b) We report the mean and standard deviation of $d_{3^\prime}$ and $d_{5^\prime}$ taken over all microstates comprising each macrostate for each sequence. The small standard deviations that are consistent with a narrow distribution of the bulk of the microstates around the free energy minimum of each macrostate. (c) Superposition of sample density for each macrostate for each sequence AT-all, GC-end, GC-core, and GC-mix projected into $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$. This panel conveys two important pieces of information. First, each macrostate is anchored around a single free energy minimum containing the preponderance of the probability mass of the microstate ensemble. These projections indicate a relatively narrowly peaked distribution of microstates around the free energy minimum with a broad tail. Second, the location of each macrostate free energy minimum in the $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$ projection readily illuminates the base pair pattern within the duplex. Within the fully hybridized H state (blue) $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 1.2 nm$^{-1}$. A 2 b.p.\ shift to 3$^\prime$ is manifest in a decrease in $d_{3^\prime}^{-1}$ (increase in $d_{3^\prime}$) and an increase in $d_{5^\prime}^{-1}$ (decrease in $d_{5^\prime}$) producing the stable 3S2 state (green). A further 2 b.p.\ shift to 3$^\prime$ results in a further decrease in $d_{3^\prime}^{-1}$ and also a decrease in $d_{5^\prime}^{-1}$ (back down to the same value as in the H state) to produce the 3S4 state (black). Analogous arguments apply to the 2 b.p.\ 5$^\prime$ shifted state 5S2 (orange) and 4 b.p.\ 5$^\prime$ shifted state 5S4 (purple). The separated dehybridized state D exists not at the origin, but, due to the finite size of our simulation box, at $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red). The GC-core frayed state F4 (orange) exists as a heart-shaped region between the H and D sates, with each symmetric lobe of the heart corresponding to fraying of one or other of the loose ends. (d) Coloring the microstates by their membership probabilities to their assigned macrostate under a fuzzy clustering procedure shows that microstates lying in the vicinity of the free energy minima have high probabilities whereas those at the macrostate cluster boundaries have less definitive assignments. The former comprise the preponderance of the microstate ensemble while the latter represent intermediate configurations that are transiently occupied as the system transitions from one macrostate to another.}}
        %\caption*{Figure S\rood{X}. Macrostate interpretation and structural analysis. The Markov state model macrostates are learned from the simulation trajectories using the featurization and deep learning procedure described in Section 2.1.2 of the main text, but it is informative to perform a \textit{post hoc} projection of the simulation data into a low-dimensional space spanned by physically interpretable order parameters in order to determine the structural characteristics of each macrostate and the structural heterogeneity of the microstate ensemble. (a) Definition of two physically-interpretable order parameters useful in characterizing the structure and base pairing of each microstate under both shifting and fraying. We define the $3^\prime$ shift for a particular base on a strand as the linear distance between that base and the base on the complementary strand offset by two base pairs in the $3^\prime$ direction. This distance reaches a minimum when the two strands are mutually shifted two base pairs in the $3^\prime$ direction. We define $d_{3^\prime}$ as the mean $3^\prime$ shift averaged over all bases in the complementary strands. It is more convenient to work the reciprocal $d_{3^\prime}^{-1}$, since it has a finite range whereas $d_{3^\prime}$ diverges for dehybridized strands. We define $d_{5^\prime}^{-1}$ analogously as the reciprocal mean $5^\prime$ shift. We have found this pair of variables extremely informative in revealing the microstate and macrostate structure and base pairing patterns. (b) Superposition of free energy surfaces for each macrostate for each sequence AT-all, GC-end, GC-core, and GC-mix projected into $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$. This panel conveys two important pieces of information. First. each macrostate is anchored around a single free energy minimum containing the preponderance of the probability mass of the microstate ensemble. Dashed lines indicate the contour around the free energy minimum containing 90\% of the microstate ensemble. These projections indicate a relatively narrowly peaked distribution of microstates around the free energy minimum with a broad tail. Second, the location of each macrostate free energy minimum in the $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$ projection readily illuminates the base pair pattern within the duplex. Within the fully hybridized H state (blue) $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 1.2 nm$^{-1}$. A 2 b.p.\ shift to 3$^\prime$ is manifest in a decrease in $d_{3^\prime}^{-1}$ (increase in $d_{3^\prime}$) and a an increase in $d_{5^\prime}^{-1}$ (increase in $d_{5^\prime}$) producing the stable 3S2 state (green). A further 2 b.p.\ shift to 3$^\prime$ results in a further decrease in $d_{3^\prime}^{-1}$ and also a decrease in $d_{5^\prime}^{-1}$ (back down to the same value as in the H state) to produce the 3S4 state (black). Analogous arguments apply to the 2 b.p.\ 5$^\prime$ shifted state 5S2 (orange) and 4 b.p.\ 5$^\prime$ shifted state 5S4 (purple). The infinitely separated dehybridized state D exists at the origin, but limited by the finite box size corresponds to $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red). The GC-core frayed state F4 (orange) exists as a heart-shaped region between the H and D sates, with each symmetric lobe of the heart corresponding to fraying of one or other of the loose ends. (c) Coloring the microstates by their membership probabilities to their assigned macrostate under a fuzzy clustering procedure shows that microstates lying in the vicinity of the free energy minima have high probabilities whereas those at the macrostate cluster boundaries have less definitive assignments. The former comprise the preponderance of the microstate ensemble while the latter represent intermediate configurations that are transiently occupied as the system transitions from one macrostate to another. (d) We report the mean and standard deviation of $d_{3^\prime}$ and $d_{5^\prime}$ taken over all microstates comprising each macrostate for each sequence. The small standard deviations that are consistent with a narrow distribution of the bulk of the microstates around the free energy minimum of each macrostate. A selection of \rood{50} microstates selected at random from those microstates with $>$90\% membership probability provide further qualitative support of a relatively homogenous structural distribution, meaning that a single high membership probability microstate is archetypal of the preponderance of microstates comprising the macrostate.}

@@

\textcolor{blue}{4)      In Figure 1 the units for rates are not provided.}

\textbf{Author Reply:}   Thank you for catching this error. Fig.~1 has been updated to include units in s$^{-1}$.

\textcolor{blue}{5)      In the comparison between experiments and theory for slow and fast T-jump IR responses (P. 17 L. 9), were there any correlations between the goodness-of-fit in the calculated rates (Figure S4) and the agreement with experiments for a given system? It is hard to judge as is because the goodness-of-fit values were not provided with Figure S4.}

\textbf{Author Reply:}     We have updated all panels of Fig.~S4 plots to include $R^2$ goodness-of-fit values determined by \rood{linear least squares fits of the models in log-space (i.e., $\log \left( f \right) = -k_d t$}) and have also re-plotted the data on log-linear axes for easier comparisons between the data and fits. The revised figure is reproduced below. In all but \rood{XX} cases we report good fits with $R^2$ $>$ \rood{0.88}. We report less good fits for \noter{which systems} likely due to \noter{why? sparser data?}, but the fits are still moderate to very good lying in the range \rood{X.XX} $<$ $R^2$ $<$ \rood{X.XX}. \noteg{Complete the placeholders here and remove all the red ink.}

\noteg{Reproduce the figure and caption here.}

@@

%\begin{figure}[ht!]
%	\begin{center}
%        \includegraphics[width=120mm]{./revision_figures/goodness_of_fit.png}
%        \caption{Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments. We have updated this supplemental figure to include goodness-of-fit values determined by the $R^2$ values of log fits and plotted both the slow and fast fits on a log scale.}
%        \label{fig:goodness-of-fit}
%	\end{center}
%\end{figure}

\textcolor{blue}{6)       The authors chose to compute the temperature jump relaxation rates from fits to quantities obtained directly from the MD simulation. However, it is not clear why the authors did not also compute the relaxation of observables from estimated Markov state models following (P. N. A. S. USA. 2011;108(12):4822-4827). This approach has also been previously used to compare kinetics obtained from Markov state models with other spectroscopic temperature-jump based experiments in DNA (J. Phys. Chem. B. 2019;123(10):2291-2304.) and RNA (J. Chem. Theory Comput. 2017;108(2):926-934) oligomers which should be mentioned. If possible, it would be interesting to compare or discuss the differences between their approach and these previous approaches.}

\textbf{Author Reply:}  The reviewer raises an important point regarding the dynamical fingerprinting methodology that have been developed to compare MSM predictions and relaxation experiments. We are familiar with these techniques, are currently applying them in our current work on abasic DNA, and did actually attempt to apply them in the present work. These analyses, however, were less informative than we had hoped due to two primary factors.

@@

First, the experimental data contains fast responses occurring on time scales of $\sim$100 ns. As detailed in Section 3.1 and Figure 1, these fast motions occur in our simulations at an acceleration factor of approximately 120$\times$ and so proceed on a time scale of $\sim$1 ns. As such, these motions lie below the $\tau$ = 1.2 ns lag time used to construct our MSMs that is required to converge the leading slow modes and produce a Markovian model capable of passing the Chapman-Kolmogorov test (cf. Figs.~S\rood{2-3}). As such, we cannot draw comparisons of our MSM against the fast experimental data.

Second, the dynamical fingerprinting method provides an elegant means to compare experimental data against an MSM constructed at the same state point. We constructed only a single MSM for each sequence at the melting temperature of the 3SPN.2 model that required the collection of 1 ms of simulation data for each sequence in order to observe sufficiently many hybridization/dehybridization events for robust fitting of the model parameters (Section 2.1.1). The experimental relaxation data, however, was collected over a range of temperatures and we wished to make comparisons across this entire range. Fitting additional MSMs was too computationally burdensome due to the high volume of data required and the cost of these simulations, so instead we conducted 120 $\times$ 1 $\mu$s simulations for each DNA sequence at each temperature. These trajectory data were too sparse to fit a robust MSM, but did allow us to make a direct comparison between the MD simulations and the experimental measurements over the entire temperature range.

@@

We therefore elected to structure our manuscript such that we first test the MD simulation predictions against experimental data in Section 3.1, then, after having validated the simulation data upon which the MSMs are constructed, perform MSM construction and analysis in Sections 3.2-6. Finally, in Section 3.7 we close with a test of a specific prediction of our MSM (the existence of long-lived metastable shifted states) that was amenable to comparison against T-jump IR.

We concur with the reviewer that we can and should explicitly mention dynamical fingerprinting approach and provide some discussion for the reader why we did not employ this approach in the present study. To this end, we have added the following text to the end of Section 3.1:

### 2021-07-29 SI_marked_up.tex
Andy edits on response letter through p. 10

\usepackage[scaled=0.85]{beramono}
\usepackage[T1]{fontenc}
\usepackage{changepage}
\usepackage{mathrsfs}
\usepackage{comment}
\usepackage[normalem]{ulem}
\usepackage{algorithm}
\usepackage{subfig}
\usepackage{algpseudocode}

@@

% Fig. S4
\begin{figure}[ht!]
	\centering
    \includegraphics[width=.6\textwidth]{FigS4.pdf}
    \caption{Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments. From 120 independent 1 $\mu$s simulations, we compiled the slow response data by recording the fraction of sequences with both central Watson-Crick base pairs intact as a function of time, and the fast response data as the fraction of sequences with both terminal Watson-Crick base pairs intact as a function of time. We define a Watson-Crick base pair to be intact if the centers-of-mass of the two comlementary bases lie within a linear distance of 1.3 nm. We extracted our computational estimate of $k_d^\mathrm{fast}$ by fitting a decaying exponential to the fraction of bound A:T termini as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$. Similarly, we extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$. \rood{We report within the figure legend to each panel the coefficient of determination $R^2$ for a least squares linear fit of the model to the data in log space (i.e., $\log \left( f \right) = -k_d t$) and data are plotted on log-linear axes to facilitate visual comparison of the fits. In all cases we observe excellent fits of the models to the data with all $R^2$ values equal to or better than 0.88 (except for the slow response at the the lowest temperature T$_m$-5K where dissociation events are sparse).}}
    \label{fig:SIFig4}
\end{figure}

@@

        \caption*{\textbf{Figure S7.}\rood{Macrostate interpretation and structural analysis. The Markov state model macrostates are learned from the simulation trajectories using the featurization and deep learning procedure described in Section 2.1.2 of the main text, but it is informative to perform a \textit{post hoc} projection of the simulation data into a low-dimensional space spanned by physically interpretable order parameters in order to determine the structural characteristics of each macrostate and the structural heterogeneity of the microstate ensemble. (a) Definition of two physically-interpretable order parameters useful in characterizing the structure and base pairing of each microstate under both shifting and fraying. We define the $3^\prime$ shift for a particular base on a strand as the linear distance between that base and the base on the complementary strand offset by two base pairs in the $3^\prime$ direction. This distance reaches a minimum when the two strands are mutually shifted two base pairs in the $3^\prime$ direction. We define $d_{3^\prime}$ as the mean $3^\prime$ shift averaged over the two central bases, represented by red arrows, in the complementary strands. It is more convenient to work the reciprocal $d_{3^\prime}^{-1}$, since it has a finite range whereas $d_{3^\prime}$ diverges for dehybridized strands. We define $d_{5^\prime}^{-1}$ analogously as the reciprocal mean $5^\prime$ shift. We have found this pair of variables extremely informative in revealing the microstate and macrostate structure and base pairing patterns. (b) We report the mean and standard deviation of $d_{3^\prime}$ and $d_{5^\prime}$ taken over all microstates comprising each macrostate for each sequence. The small standard deviations that are consistent with a narrow distribution of the bulk of the microstates around the free energy minimum of each macrostate. (c) Superposition of sample density for each macrostate for each sequence AT-all, GC-end, GC-core, and GC-mix projected into $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$. This panel conveys two important pieces of information. First. each macrostate is anchored around a single free energy minimum containing the preponderance of the probability mass of the microstate ensemble. These projections indicate a relatively narrowly peaked distribution of microstates around the free energy minimum with a broad tail. Second, the location of each macrostate free energy minimum in the $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$ projection readily illuminates the base pair pattern within the duplex. Within the fully hybridized H state (blue) $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 1.2 nm$^{-1}$. A 2 b.p.\ shift to 3$^\prime$ is manifest in a decrease in $d_{3^\prime}^{-1}$ (increase in $d_{3^\prime}$) and a an increase in $d_{5^\prime}^{-1}$ (increase in $d_{5^\prime}$) producing the stable 3S2 state (green). A further 2 b.p.\ shift to 3$^\prime$ results in a further decrease in $d_{3^\prime}^{-1}$ and also a decrease in $d_{5^\prime}^{-1}$ (back down to the same value as in the H state) to produce the 3S4 state (black). Analogous arguments apply to the 2 b.p.\ 5$^\prime$ shifted state 5S2 (orange) and 4 b.p.\ 5$^\prime$ shifted state 5S4 (purple). The infinitely separated dehybridized state D exists at the origin, but limited by the finite box size corresponds to $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red). The GC-core frayed state F4 (orange) exists as a heart-shaped region between the H and D sates, with each symmetric lobe of the heart corresponding to fraying of one or other of the loose ends. (d) Coloring the microstates by their membership probabilities to their assigned macrostate under a fuzzy clustering procedure shows that microstates lying in the vicinity of the free energy minima have high probabilities whereas those at the macrostate cluster boundaries have less definitive assignments. The former comprise the preponderance of the microstate ensemble while the latter represent intermediate configurations that are transiently occupied as the system transitions from one macrostate to another.}}

>>>>

\usepackage[scaled=0.85]{beramono}
\usepackage[T1]{fontenc}
\usepackage{changepage}
\usepackage{mathrsfs}
\usepackage{comment}
\usepackage[normalem]{ulem}
\usepackage{algorithm}
\usepackage{subfig}
\usepackage{algpseudocode}
\usepackage[font={small,sf}, singlelinecheck=false]{caption}
\usepackage{capt-of}

@@

% Fig. S4
\begin{figure}[ht!]
	\centering
	\includegraphics[width=.97\textwidth]{FigS4.pdf}
\end{figure}

\begin{figure}[ht!]
	\begin{center}
    	\caption{Exponential fits for both slow (dissociation) and fast (fraying) response for all four sequences during ``computational T-jump'' experiments. From 120 independent 1 $\mu$s simulations, we compiled the slow response data by recording the fraction of sequences with both central Watson-Crick base pairs intact as a function of time, and the fast response data as the fraction of sequences with both terminal Watson-Crick base pairs intact as a function of time. We define a Watson-Crick base pair to be intact if the centers-of-mass of the two comlementary bases lie within a linear distance of 1.3 nm. We extracted our computational estimate of $k_d^\mathrm{fast}$ by fitting a decaying exponential to the fraction of bound A:T termini as a function of time $f_\mathrm{unfrayed}(t) = \exp(-k_d^\mathrm{fast}t)$. Similarly, we extracted our computational estimate of $k_d^\mathrm{slow}$ by fitting a decaying exponential to the fraction of hybridized sequences as a function of time $f_\mathrm{hybridized}(t) = \exp(-k_d^\mathrm{slow}t)$. \rood{We report within the figure legend to each panel the coefficient of determination $R^2$ for a least squares linear fit of the model to the data in log space (i.e., $\log \left( f \right) = -k_d t$) and data are plotted on log-linear axes to facilitate visual comparison of the fits. In all cases we observe excellent fits of the models to the data with all $R^2$ values equal to or better than 0.88 (except for the slow response at the the lowest temperature T$_m$-5K where dissociation events are sparse).}}
    \label{fig:SIFig4}
    	\end{center}
\end{figure}

@@

        \caption*{Figure S7. \rood{Macrostate interpretation and structural analysis. The Markov state model macrostates are learned from the simulation trajectories using the featurization and deep learning procedure described in Section 2.1.2 of the main text, but it is informative to perform a \textit{post hoc} projection of the simulation data into a low-dimensional space spanned by physically interpretable order parameters in order to determine the structural characteristics of each macrostate and the structural heterogeneity of the microstate ensemble. (a) Definition of two physically-interpretable order parameters useful in characterizing the structure and base pairing of each microstate under both shifting and fraying. We define the $3^\prime$ shift for a particular base on a strand as the linear distance between that base and the base on the complementary strand offset by two base pairs in the $3^\prime$ direction. This distance reaches a minimum when the two strands are mutually shifted two base pairs in the $3^\prime$ direction. We define $d_{3^\prime}$ as the mean $3^\prime$ shift averaged over the two central bases, represented by red arrows, in the complementary strands. It is more convenient to work the reciprocal $d_{3^\prime}^{-1}$, since it has a finite range whereas $d_{3^\prime}$ diverges for dehybridized strands. We define $d_{5^\prime}^{-1}$ analogously as the reciprocal mean $5^\prime$ shift. We have found this pair of variables extremely informative in revealing the microstate and macrostate structure and base pairing patterns. (b) We report the mean and standard deviation of $d_{3^\prime}$ and $d_{5^\prime}$ taken over all microstates comprising each macrostate for each sequence. The small standard deviations that are consistent with a narrow distribution of the bulk of the microstates around the free energy minimum of each macrostate. (c) Superposition of sample density for each macrostate for each sequence AT-all, GC-end, GC-core, and GC-mix projected into $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$. This panel conveys two important pieces of information. First, each macrostate is anchored around a single free energy minimum containing the preponderance of the probability mass of the microstate ensemble. These projections indicate a relatively narrowly peaked distribution of microstates around the free energy minimum with a broad tail. Second, the location of each macrostate free energy minimum in the $d_{3^\prime}^{-1}$-$d_{5^\prime}^{-1}$ projection readily illuminates the base pair pattern within the duplex. Within the fully hybridized H state (blue) $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 1.2 nm$^{-1}$. A 2 b.p.\ shift to 3$^\prime$ is manifest in a decrease in $d_{3^\prime}^{-1}$ (increase in $d_{3^\prime}$) and an increase in $d_{5^\prime}^{-1}$ (decrease in $d_{5^\prime}$) producing the stable 3S2 state (green). A further 2 b.p.\ shift to 3$^\prime$ results in a further decrease in $d_{3^\prime}^{-1}$ and also a decrease in $d_{5^\prime}^{-1}$ (back down to the same value as in the H state) to produce the 3S4 state (black). Analogous arguments apply to the 2 b.p.\ 5$^\prime$ shifted state 5S2 (orange) and 4 b.p.\ 5$^\prime$ shifted state 5S4 (purple). The separated dehybridized state D exists not at the origin, but, due to the finite size of our simulation box, at $d_{3^\prime}^{-1}$ $\approx$ $d_{5^\prime}^{-1}$ $\approx$ 0.2 nm$^{-1}$ (red). The GC-core frayed state F4 (orange) exists as a heart-shaped region between the H and D sates, with each symmetric lobe of the heart corresponding to fraying of one or other of the loose ends. (d) Coloring the microstates by their membership probabilities to their assigned macrostate under a fuzzy clustering procedure shows that microstates lying in the vicinity of the free energy minima have high probabilities whereas those at the macrostate cluster boundaries have less definitive assignments. The former comprise the preponderance of the microstate ensemble while the latter represent intermediate configurations that are transiently occupied as the system transitions from one macrostate to another.}}

### 2021-07-29 manuscript_marked_up.tex
Andy edits on response letter through p. 10

%\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}
%\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes 
%\newcommand*{\noter}[1]{}					% notes off
%\newcommand*{\noteb}[1]{\textcolor{blue}{[[#1]]}}		% notes

@@

In Fig.~\ref{fig:relaxation-comparison}a we observe an exponential increase of $k_d^\mathrm{slow}$ with temperature, as expected from the large enthalpic barrier to duplex dissociation.\cite{Craig1971RelaxationOligonucleotides,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC} Under the time and temperature calibration corrections, we see generally very good agreement between the computational and experimental curves. Of the four sequences, GC-core shows the largest discrepancy between computation and experiment, although the general exponential trend is preserved. This may be a result of its high propensity to fray (cf.\ Section \ref{sec:GC-core}). In Fig.~\ref{fig:relaxation-comparison}b we expose a largely linear dependence of $k_d^\mathrm{fast}$ upon temperature for AT-all, GC-end, and GC-mix compared to an exponential dependence for GC-end. These trends can be understood in light of the comparatively larger enthalpic barrier for dissociation of the terminal G:C base pair in GC-end compared to that for the A:T terminal pair in the other three sequences. Again we see good agreement between the scaled computational predictions and the experimental T-jump measurements. The favorable comparison of computation and experiment provides support for the capacity of the 3SPN.2 model to reliably reproduce sequence-dependent trends in the slow and fast kinetics of the four DNA oligomers \rood{, and lends confidence in the use of these data for the parameterization of Markov state models of the long-time dynamics of each sequence. We observe that the dynamical fingerprinting approach presents an elegant means to compare experimental relaxation data directly against a Markov state model extracted from simulation data in terms of a ``fingerprint'' of peaks with amplitudes and time scales related to the relaxation of particular system observables~\citep{Noe2011DynamicalExperiments}. These techniques have been previously employed to in applications to base stacking of DNA dinucleoside monophosphates~\citep{Remington2019MolecularKinetics} and RNA~\citep{Pinamonti2017}. We explored the use of this approach to validate the MD simulation data but encountered challenges in resolving fast dynamical motions below the lag time of the fitted MSM and the extremely high computational cost of collecting sufficient simulation data to fit MSMs at each temperature for which experimental data was collected. Accordingly, we instead elected to perform a direct comparison between the experimental data and MD trajectories to validate the simulations themselves, then proceed to train MSMs over these data and conduct analysis and experimental tests of the MSM predictions.}.

>>>>

\newcommand*{\addref}[1]{\textcolor{red}{\{ADD REF: #1\}}}
\newcommand*{\noter}[1]{\textcolor{red}{[[#1]]}}		% notes 
%\newcommand*{\noter}[1]{}					% notes off
\newcommand*{\noteb}[1]{\textcolor{blue}{[[#1]]}}		% notes

@@

In Fig.~\ref{fig:relaxation-comparison}a we observe an exponential increase of $k_d^\mathrm{slow}$ with temperature, as expected from the large enthalpic barrier to duplex dissociation.\cite{Craig1971RelaxationOligonucleotides,Porschke1971CooperativeTransition,Williams1989LaserDGCATGC} Under the time and temperature calibration corrections, we see generally very good agreement between the computational and experimental curves. Of the four sequences, GC-core shows the largest discrepancy between computation and experiment, although the general exponential trend is preserved. This may be a result of its high propensity to fray (cf.\ Section \ref{sec:GC-core}). In Fig.~\ref{fig:relaxation-comparison}b we expose a largely linear dependence of $k_d^\mathrm{fast}$ upon temperature for AT-all, GC-end, and GC-mix compared to an exponential dependence for GC-end. These trends can be understood in light of the comparatively larger enthalpic barrier for dissociation of the terminal G:C base pair in GC-end compared to that for the A:T terminal pair in the other three sequences. Again we see good agreement between the scaled computational predictions and the experimental T-jump measurements. The favorable comparison of computation and experiment provides support for the capacity of the 3SPN.2 model to reliably reproduce sequence-dependent trends in the slow and fast kinetics of the four DNA oligomers\rood{, and lends confidence in the use of these data for the parameterization of Markov state models of the long-time dynamics of each sequence. We observe that the dynamical fingerprinting approach presents an elegant means to compare experimental relaxation data directly against a Markov state model extracted from simulation data in terms of a ``fingerprint'' of peaks with amplitudes and time scales related to the relaxation of particular system observables~\citep{Noe2011DynamicalExperiments}. These techniques have been previously employed to in applications to base stacking of DNA dinucleoside monophosphates~\citep{Remington2019MolecularKinetics} and RNA~\citep{Pinamonti2017}. We explored the use of this approach to validate the MD simulation data but encountered challenges in resolving fast dynamical motions below the lag time of the fitted MSM and the extremely high computational cost of collecting sufficient simulation data to fit MSMs at each temperature for which experimental data was collected. Accordingly, we instead elected to perform a direct comparison between the experimental data and MD trajectories to validate the simulations themselves, then proceed to train MSMs over these data and conduct analysis and experimental tests of the MSM predictions.}.
